Factors Shaping Financial Well-Being: Examining Demographics, Psychological Elements, and Financial Knowledge

### 2024-12-04 paper/paper.qmd
Update paper for prose concerns

---
title: "Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy"
subtitle: "Psychological Distress Has the Strongest Negative Impact on Financial Outcomes"
author: "Xinxiang Gao"
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
abstract: "This paper analyzes the determinants of financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey. Using a comprehensive regression analysis of over 6,000 respondents, we find that psychological distress has the strongest negative association with financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale. Higher income and age are associated with better financial outcomes, while education shows modest positive effects. These findings suggest that financial education programs should incorporate mental health support, particularly for vulnerable populations."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

# Introduction

In today's complex financial landscape, achieving financial well-being is a fundamental goal for individuals and households. While financial well-being encompasses the ability to meet current and future financial obligations, its determinants span multiple domains including financial literacy, demographic characteristics, and psychosocial factors. Understanding these relationships is crucial for developing effective policies and interventions to improve financial outcomes across diverse population segments.

To measure financial well-being systematically, the Consumer Financial Protection Bureau (CFPB) developed the Financial Well-Being Scale (FWBscore), a 10-item questionnaire assessing four critical elements: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and ability to make life-improving choices. Alongside this measure, researchers have developed various instruments to assess financial knowledge and skills. The Lusardi and Mitchell financial knowledge scale (LMscore) by @lusardi2008 provides a foundational three-item measure of basic financial concepts, while the Knoll and Houts financial knowledge scale (KHscore) by @knoll2012 offers a more comprehensive 10-item assessment using item response theory. Additionally, the CFPB Financial Skill Scale (FSscore) by @cfpb2015 evaluates practical financial competencies through an IRT-based approach.

While previous research has examined financial literacy's role in economic outcomes, less attention has been paid to how demographic and psychosocial factors interact with financial knowledge to influence overall financial well-being. This gap is particularly significant given evidence suggesting that financial knowledge alone may not fully explain variations in financial outcomes across different population segments.

Our primary estimand is the average change in financial well-being score associated with variations in demographic and psychosocial characteristics, holding other factors constant. This causal parameter helps identify which factors have the strongest relationship with financial well-being, enabling targeted policy interventions.

Our findings indicate that psychological distress has the strongest negative association with financial well-being (-16.5 points), while higher income ($150,000+) shows the largest positive effect (+12.6 points). Advanced age (75+ years) and excellent health status are also strong positive predictors (+9.4 and +7.7 points respectively). Notably, financial knowledge measures show relatively modest effects compared to these psychological and demographic factors, suggesting a need to rethink traditional approaches to financial education.

The rest of the paper is organized as follows: @sec-data describes our data and methodology. @sec-model presents our modeling approach. @sec-results presents our empirical results. @sec-discussion discusses the implications of our findings and concludes. @sec-app provides additional details on the survey methodology and @sec-ack provides acknowledgements.

@@

The dataset for this study is the National Financial Well-Being Survey Public Use File (PUF) collected by @cfpb_fwb_survey. This dataset is a rich resource that includes 217 variables encompassing financial, demographic, and psychosocial factors, drawn from a representative sample of the U.S. population. It provides data on key indicators such as financial knowledge, skills, behaviors, and overall well-being. These data are ideal for analyzing relationships between financial well-being and other financial or non-financial factors.

The survey was conducted using the GfK KnowledgePanel, an online probability-based panel designed to be representative of the U.S. adult population. The GfK KnowledgePanel is created by randomly selecting households and inviting them to participate in the panel. Households without internet access are provided with a computer and internet service. This method ensures that the sample is representative across various demographic groups in the United States. Additionally, the dataset incorporates external contextual variables, such as poverty levels in respondents’ counties of residence, offering a nuanced perspective on financial well-being in diverse settings. While other datasets, such as the Survey of Consumer Finances or the Panel Study of Income Dynamics, also include financial information, they do not provide a standardized, validated measure of financial well-being like the FWBscore. This makes the CFPB dataset uniquely suited for this study.

@@

### Primary Outcome Variable

Financial Well-Being Score (FWBscore): The CFPB Financial Well-Being Scale score is the key outcome variable. It is a 10-item scale capturing four dimensions: control over day-to-day finances, capacity to absorb financial shocks, progress toward financial goals, and freedom of financial choices. Scores range from 0 to 100, with higher scores indicating greater financial well-being.

@@

All the independent variables are collected through self-reported questionnaire responses, as addressed below.

Age (agecat): Categorized into age groups (e.g., 18–24, 25–34, 35–44, 45–54, 55–64, 65+).

Education (PPEDUC): Highest degree attained, categorized from less than high school to graduate degrees.

Household Income (PPINCIMP): Categorized income levels (e.g., <$20,000, $20,000-$39,999, $40,000-$59,999, etc.).

Marital Status (PPMARIT): Categories include single, married, divorced, widowed, and other.

Health (HEALTH): Self-reported general health, rated on a 5-point Likert scale from 1 (Poor) to 5 (Excellent).

Stress (DISTRESS): Perceived level of stress, rated on a 5-point Likert scale from 1 (Not at all) to 5 (Extremely).

@@

Lusardi and Mitchell Financial Knowledge Scale Score (LMscore): A 3-item summative scale measuring financial literacy, particularly in planning and decision-making.

Knoll and Houts Financial Knowledge Scale Score (KHscore): A 10-item IRT-based measure assessing broader financial knowledge across multiple dimensions.

Financial Skill Scale Score (FSscore): An IRT-based measure developed by the CFPB, capturing practical financial skills like budgeting, debt management, and saving.

@@

@fig-histograms shows the distribution of four different financial scores (FSscore, FWBscore, KHscore, and LMscore) through histograms. FSscore and FWBscore show continuous distributions with bell-shaped curves, though not perfectly normal: FSscore ranges from about 0-80 with a peak around 40-50, while FWBscore has a similar range but appears more symmetrically distributed with its peak around 45-50. The KHscore shows a discrete distribution across four main categories (from -2 to 1), with the highest frequency at 0 and relatively balanced frequencies between -1 and 1. The LMscore is also discrete but shows a strong positive skew, with values ranging from 0 to 3 and a clear majority of responses at the highest value (3). This pattern suggests that while financial wellbeing and financial situations (FWB and FS) vary normally across the population, the literacy measure (LM) indicates most respondents have high literacy levels.

@@

The correlation matrix heatmap as @fig-correlations shows the relationships between four different scores: FWBscore, FSscore, LMscore, and KHscore. The strongest correlation appears between LMscore and KHscore with a moderate positive correlation of 0.6, followed by a moderate correlation of 0.49 between FWBscore and FSscore. FWBscore shows weak to moderate positive correlations with LMscore (0.23) and KHscore (0.33), while FSscore exhibits the weakest correlations with LMscore (0.11) and KHscore (0.19). All correlations are positive, suggesting that as one score increases, the others tend to increase as well, though with varying strengths of association. The diagonal shows perfect correlations of 1.0, as expected when comparing a variable with itself.

@@

@fig-barplots provides a comprehensive demographic overview of the survey population. The age distribution shows strong representation across middle-age groups (25-34, 45-54, and 62-69) with fewer respondents in the youngest (18-24) and oldest (70-74) categories. Education levels indicate that "Some college/Associate" and "High school degree/GED" are the most common, while "Less than high school" is the least common. Income distribution is relatively balanced across categories, with slightly higher numbers in the $100,000-$149,999 range. Marital status shows a clear majority of "Married" respondents, significantly outnumbering other categories. Health status data reveals most respondents report "Very good" or "Good" health, with few reporting "Poor" health. Finally, psychological distress measurements show a normal-like distribution centered around "Neither agree nor disagree" and "Agree" responses, with fewer responses at the extremes ("Strongly agree" and "Strongly disagree").

@@

We employ two complementary regression models to analyze financial well-being. Our primary model examines the relationship between financial well-being and demographic and psychosocial factors, while a supplementary model focuses on the role of financial knowledge and skills.

## Primary Model

The primary model can be expressed as:

$$
\text{FWB}_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Education}_i + \beta_3 \text{Income}_i+ \beta_4 \text{Marital}_i + \beta_5 \text{Health}_i + \beta_6 \text{Distress}_i + \epsilon_i
$$

where $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$

Here:

- $\text{FWB}_i$ is the Financial Well-Being score for individual $i$;

- Age categories are represented by indicator variables with 18-24 as the baseline;

- Education levels are coded with "Less than high school" as the baseline;

- Income categories start from "Less than $20,000" as the baseline;

- Marital status uses "Married" as the reference category;

- Health status is measured from "Poor" (baseline) to "Excellent";

- Psychological distress ranges from "Strongly disagree" (baseline) to "Strongly agree".

## Supplementary Model

The supplementary model examining financial literacy's role is:

$$
\text{FWB}_i = \beta_0 + \beta_1 \text{LMscore}_i + \beta_2 \text{KHscore}_i + \beta_3 \text{FSscore}_i + \epsilon_i
$$

where the scores represent standardized measures of financial knowledge and skills.

@@

We conducted several diagnostic checks to validate our model assumptions, as illustrated in @fig-model-diagnostics. The residual plots show relatively homoscedastic errors and no serious violations of linearity. The Q-Q plot indicates approximately normal residuals, though with some deviation in the tails. The Cook's distance plot identifies no highly influential observations that might distort our results.

@@

Our analysis reveals several key patterns in the determinants of financial well-being. The coefficient table @tbl-output visualizes the main findings from our primary model, showing the relative importance of different factors.

## Age Effects
Age demonstrates a strong positive relationship with financial well-being, particularly for older adults. Compared to the 18-24 age group, individuals aged 75+ show the largest positive effect (+9.37 points), followed by those 70-74 (+7.51 points) and 62-69 (+6.35 points). This suggests that financial well-being tends to improve substantially with age, possibly reflecting accumulated wealth and experience.

## Income and Education
Income exhibits the strongest positive association with financial well-being among all variables. Those earning $150,000 or more score 12.60 points higher than those earning under $20,000, with a clear gradient across income categories. Higher education also shows positive effects, with graduate degree holders scoring 2.92 points higher than those without high school completion.

## Health and Psychological Distress
Health status shows a strong positive relationship with financial well-being, while psychological distress demonstrates the largest negative effects in our model. Excellent health is associated with a 7.70-point increase compared to poor health. Conversely, high psychological distress ("Strongly agree") corresponds to a substantial 16.52-point decrease in financial well-being.

## Financial Knowledge and Skills
Our supplementary model, as in Table 2, requires careful interpretation given the distinct distributions of our financial literacy measures:

Financial skills (FSscore) shows a moderate positive effect ($\beta$ = 0.501), which is meaningful given its approximately normal distribution across the population.

While Knoll and Houts financial knowledge (KHscore) shows a large coefficient ($\beta$ = 3.748), this should be interpreted with caution given its discrete, roughly uniform distribution across four categories (-2 to 1).

The Lusardi and Mitchell financial knowledge (LMscore) effect ($\beta$ = 0.874) must be interpreted carefully due to its highly skewed distribution, with most respondents scoring at the maximum value of 3.

These results suggest that while all measures show positive associations with financial well-being, their interpretations need to account for their distinct distributional characteristics. The FSscore's effect may be the most reliably interpretable given its normal distribution, while the strong coefficient for KHscore and moderate coefficient for LMscore should be considered in light of their non-normal distributions. These distributional patterns suggest that future research might benefit from more granular measures of financial knowledge that could better capture variability across the population.

@@

## Overview of Research Contribution

This paper provides a comprehensive analysis of how psychological factors, particularly mental distress, interact with traditional socioeconomic determinants to influence financial well-being. Using the CFPB's National Financial Well-Being Survey, we demonstrate that psychological distress has a larger negative association with financial well-being than previously recognized determinants such as income, education, or financial literacy. This finding challenges the conventional focus on financial education as the primary lever for improving financial outcomes.

## Key Insights About Financial Well-Being

### The Primacy of Psychological Factors

Our most striking finding is the dominant role of psychological distress in determining financial well-being. A one-point increase in psychological distress is associated with a 4.89-point decrease in financial well-being—an effect nearly three times larger than that of a year of education. This suggests that mental health may be a crucial prerequisite for financial success, potentially explaining why some financially literate individuals still struggle with financial management. The magnitude of this relationship indicates that mental health support could be as important as financial education in promoting financial well-being.

### The Limited Role of Financial Knowledge

Our second key insight concerns the relatively modest impact of financial knowledge on financial well-being. While all three measures of financial knowledge show positive associations with financial well-being, their effects are smaller than those of psychological and demographic factors. This finding suggests that the traditional policy emphasis on financial education may need to be reconsidered. Financial knowledge appears necessary but not sufficient for achieving financial well-being, particularly when individuals face psychological challenges.

## Study Limitations

Several limitations of our analysis warrant discussion. First, our cross-sectional data cannot establish causal relationships. The association between psychological distress and financial well-being likely reflects bidirectional causality—poor financial circumstances may increase distress, while distress may impair financial decision-making. Future research using longitudinal data could help disentangle these effects.

Second, our measures of psychological distress and financial well-being rely on self-reported data, which may be subject to various biases. Individuals experiencing financial difficulties might be more likely to report psychological distress, potentially amplifying the observed relationship between these variables.

Third, while our sample is nationally representative, it may not fully capture the experiences of particularly vulnerable populations, such as the unbanked or those without internet access. These groups might face unique challenges that our analysis cannot address.

## Policy Implications and Future Directions

Our findings have important implications for policy and practice. First, financial wellness programs should consider incorporating mental health support or at least screening for psychological distress. The strong negative association between distress and financial well-being suggests that addressing mental health challenges could be a prerequisite for effective financial education.

Second, policymakers should consider broadening the scope of financial capability initiatives beyond traditional education. Programs that build psychological resilience and stress management skills might be as important as those teaching financial concepts. This could be particularly relevant for vulnerable populations who face multiple stressors.

Third, financial institutions and counselors might benefit from training in recognizing signs of psychological distress and making appropriate referrals to mental health resources. This could help create a more holistic approach to promoting financial well-being.

Future research in financial well-being should prioritize three key directions. First, longitudinal studies are needed to establish clear causal relationships between psychological distress and financial outcomes, moving beyond the limitations of cross-sectional data. Second, researchers should focus on intervention studies that evaluate integrated programs combining financial education with mental health support, as this could provide practical evidence for policy design. Finally, understanding demographic variations in how psychological distress affects financial well-being across different population groups would help tailor interventions to those most in need.

In conclusion, our findings suggest that achieving financial well-being requires more than just financial knowledge—it demands attention to psychological well-being as well. This insight calls for a fundamental rethinking of how we approach financial education and support, pointing toward more integrated approaches that address both the financial and psychological aspects of economic decision-making.

@@

The CFPB Financial Well-Being Scale methodology merits critical examination across three key dimensions: sampling design, measurement validation, and methodological limitations.

### Sampling Framework and Implementation

The survey employed a multi-round data collection approach through a large nationally recognized provider of survey data. The initial round fielded 46 candidate items to 4,500 respondents, with the sample divided between adults aged 18 to 61 and those aged 62 and older. This approach aimed for a diverse consumer base, though not necessarily nationally representative, ensuring that the sample was roughly similar to the U.S. population based on Census demographics for working-age and older consumers. The inclusion of both younger and older age groups, as well as varying modes of administration (self-administered and interviewer-administered), is a strength in addressing potential mode effects and providing insights into how different populations respond to the scale.

However, despite efforts to achieve diversity, there are inherent challenges in fully representing transient or hard-to-reach populations, which could introduce selection bias. Moreover, the exclusion of approximately 500 pencil-and-paper surveys due to lack of confidence in these data suggests variability in data quality across different collection methods, potentially affecting the generalizability of findings.

### Measurement Development and Validation

The development of the Financial Well-Being Scale adhered to item response theory (IRT) principles, involving extensive rounds of refinement from an initial pool of items down to the final set. The use of exploratory and confirmatory factor analysis, alongside IRT modeling, provided robust evidence for the scale's structure. Psychometric analyses conducted by Vector Psychometric Group supported the use of a bi-factor model, capturing an overall financial well-being score while accounting for positive and negative wording effects within items.

The reported marginal reliability of at least 0.89 indicates strong internal consistency, and the abbreviated version also showed acceptable reliability above 0.80. However, the absence of temporal stability metrics limits the understanding of the scale's consistency over time. Furthermore, the reliance on cognitive interviews with a relatively small sample size (n=19) may limit the breadth of perspectives captured, especially across diverse populations.

Validation of the scale included testing its performance across different demographic groups and survey modes, supporting the comparability of scores derived from different contexts. Known-groups validation helped establish discriminant validity, and correlations with objective financial measures provided criterion validity. Nevertheless, cross-validation with external datasets remains limited, which can affect the extent to which the results can be generalized beyond the study sample.

### Methodological Strengths and Limitations

The mixed-mode implementation allowed for the assessment of mode effects, which is important given documented differences in self-reporting behaviors depending on the survey format. Quality control measures such as attention checks and monitoring of completion times were used to ensure data integrity, although the specific criteria applied and their impact on the final sample composition are not detailed.

Critical limitations include:
1. Potential underrepresentation of certain populations despite the address-based sampling methodology.
2. Self-report bias in financial measures, particularly relevant for sensitive topics.
3. Limited assessment of the temporal stability of the financial well-being construct.

The study achieved a response rate that exceeds typical online panel rates, but this also raises questions about nonresponse bias. While multiple contact attempts and incentives were utilized, a deeper analysis of nonresponse patterns would enhance methodological transparency.

In summary, the methodological choices reflect a balance between practical constraints and ideal survey design standards. The approach aligns with contemporary practices in survey research while acknowledging limitations common in large-scale financial studies.

>>>>

---
title: "Factors Shaping Financial Well-Being: Examining Demographics, Psychological Elements, and Financial Knowledge"
subtitle: "Psychological Distress and Financial Health: Measuring the Relationship"
author: "Xinxiang Gao"
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
abstract: "This paper examines factors affecting financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey. Through regression analysis of over 6,000 respondents, we found that psychological distress showed the strongest negative connection to financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale. People with higher income and those in older age groups reported better financial outcomes, while education demonstrated small positive effects. These results indicate that financial education programs would benefit from including mental health support, especially for people facing financial challenges."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

# Introduction
Financial well-being forms a core objective for individuals and households navigating modern economic conditions. The concept encompasses meeting present and future financial obligations, influenced by financial literacy, demographics, and psychological factors. Understanding these relationships informs policy design and programs to enhance financial outcomes across populations.

The Consumer Financial Protection Bureau (CFPB) created the Financial Well-Being Scale (FWBscore), a 10-item questionnaire measuring four components: control over daily finances, ability to handle financial emergencies, progress on financial objectives, and freedom to make life choices. Researchers have developed several tools to evaluate financial knowledge. The Lusardi and Mitchell financial knowledge scale (LMscore) by @lusardi2008 uses three items to measure basic financial concepts. The Knoll and Houts financial knowledge scale (KHscore) by @knoll2012 employs a 10-item assessment based on item response theory. The CFPB Financial Skill Scale (FSscore) by @cfpb2015 measures practical financial abilities using IRT methods.

Research has studied how financial literacy affects economic outcomes. However, questions remain about how demographics and psychological factors combine with financial knowledge to shape financial well-being. Understanding these relationships could better explain differences in financial outcomes across population groups.

Our study examines how changes in demographic and psychological characteristics relate to financial well-being scores, while controlling for other variables. This analysis identifies which factors most strongly influence financial well-being, helping guide policy decisions.

The results show psychological distress has the strongest negative link to financial well-being (-16.5 points). Higher income ($150,000+) shows the largest positive relationship (+12.6 points). Age over 75 years and excellent health status also predict better outcomes (+9.4 and +7.7 points respectively). Financial knowledge measures show smaller effects than these psychological and demographic factors, suggesting new approaches to financial education may be needed.

The paper continues as follows: @sec-data describes our data and methods. @sec-model explains our modeling approach. @sec-results presents our findings. @sec-discussion examines implications and concludes. @sec-app details survey methods and @sec-ack lists acknowledgements.

@@

The National Financial Well-Being Survey Public Use File (PUF) by @cfpb_fwb_survey provides the data for this study. The dataset contains 217 variables measuring financial, demographic, and psychological factors from a representative U.S. population sample. It records financial knowledge, skills, behaviors, and well-being.

GfK KnowledgePanel collected the survey data using an online probability-based panel representing U.S. adults. The panel forms through random household selection and invitation. GfK provides computers and internet service to households lacking access, ensuring representation across U.S. demographic groups. The dataset includes contextual variables like county-level poverty rates, providing environmental context for financial well-being assessments. While other datasets like the Survey of Consumer Finances and Panel Study of Income Dynamics contain financial information, they lack the standardized FWBscore measure, making this CFPB dataset particularly suitable.

@@

### Primary Outcome Variable
Financial Well-Being Score (FWBscore): This 10-item CFPB scale measures four areas: control over day-to-day finances, capacity to handle financial emergencies, progress toward financial goals, and freedom of financial choices. Scores range from 0 to 100, with higher scores showing better financial well-being.

@@

The data comes from self-reported questionnaire responses:

Age (agecat): Groups include 18–24, 25–34, 35–44, 45–54, 55–64, and 65+.

Education (PPEDUC): Categories range from less than high school to graduate degrees.

Household Income (PPINCIMP): Groups span from <$20,000 to higher brackets like $20,000-$39,999 and $40,000-$59,999.

Marital Status (PPMARIT): Includes single, married, divorced, widowed, and other.

Health (HEALTH): Self-reported rating from 1 (Poor) to 5 (Excellent).

Stress (DISTRESS): Self-reported rating from 1 (Not at all) to 5 (Extremely).

Using categories for continuous variables like age and income helps interpret results across demographic groups, enables statistical group comparisons, and matches policy planning needs. This grouping also protects respondent privacy while maintaining analytical detail. The standard categories allow comparison with other demographic research.

@@

Lusardi and Mitchell Financial Knowledge Scale Score (LMscore): Three items measuring financial literacy in planning and decision-making.

Knoll and Houts Financial Knowledge Scale Score (KHscore): Ten items using IRT methods to assess financial knowledge.

Financial Skill Scale Score (FSscore): CFPB-developed IRT measure of practical financial skills including budgeting, debt management, and saving.

@@

@fig-histograms displays distributions of four financial measures: FSscore, FWBscore, KHscore, and LMscore. The FSscore and FWBscore follow roughly bell-shaped patterns. FSscore spans 0-80, peaking between 40-50, while FWBscore shows a more balanced distribution centered at 45-50. KHscore plots across four points (-2 to 1), with the highest frequency at 0 and similar frequencies at -1 and 1. LMscore ranges from 0 to 3, with most responses concentrated at 3. These patterns indicate that financial well-being and financial situations distribute evenly through the population, while most respondents score high on basic financial literacy.

@@

@fig-correlations presents a correlation matrix heatmap comparing FWBscore, FSscore, LMscore, and KHscore. LMscore and KHscore show the strongest relationship (0.6), followed by FWBscore and FSscore (0.49). FWBscore pairs with LMscore (0.23) and KHscore (0.33), showing smaller positive relationships. FSscore has the smallest relationships with LMscore (0.11) and KHscore (0.19). The positive values indicate that scores tend to move in the same direction. Each measure's correlation with itself equals 1.0 on the diagonal.

@@

@fig-barplots presents the demographic patterns of survey participants. Age groups 25-34, 45-54, and 62-69 have the highest numbers of respondents, while age groups 18-24 and 70-74 show lower participation. In education, "Some college/Associate" and "High school degree/GED" form the largest groups, with "Less than high school" having the smallest representation. Income groups distribute evenly, with a slight peak in the $100,000-$149,999 range. Married respondents make up the largest portion of marital status categories. For health status, most people report "Very good" or "Good" health, with few selecting "Poor." Psychological distress responses center between "Neither agree nor disagree" and "Agree," tapering at both ends.

@@

We use two regression models to study financial well-being: one examining demographic and psychological factors, and another focusing on financial knowledge and skills.

## Primary Model
The primary model takes this form:
$$
\text{FWB}_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Education}_i + \beta_3 \text{Income}_i+ \beta_4 \text{Marital}_i + \beta_5 \text{Health}_i + \beta_6 \text{Distress}_i + \epsilon_i
$$
where $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$

The variables represent:
- $\text{FWB}_i$: Financial Well-Being score for individual $i$
- Age: Categories with 18-24 as baseline
- Education: Levels starting at "Less than high school" as baseline
- Income: Categories beginning with "Less than $20,000" as baseline
- Marital status: Using "Married" as baseline
- Health status: From "Poor" (baseline) to "Excellent"
- Psychological distress: From "Strongly disagree" (baseline) to "Strongly agree"

## Supplementary Model
This model examines how financial literacy relates to well-being:
$$
\text{FWB}_i = \beta_0 + \beta_1 \text{LMscore}_i + \beta_2 \text{KHscore}_i + \beta_3 \text{FSscore}_i + \epsilon_i
$$
where each score represents a standardized measure of financial knowledge and skills.

@@

The model checks in @fig-model-diagnostics support our statistical assumptions. The residual plots show consistent error variance and maintain linear patterns. The Q-Q plot confirms mostly normal residuals, with slight tail variations. The Cook's distance measurements indicate no single observation unduly affects our results.

@@

Our analysis identifies clear patterns in financial well-being factors. Table @tbl-output presents the primary model results.

## Age Effects
Older adults report higher financial well-being compared to those aged 18-24. People aged 75+ score 9.37 points higher, followed by ages 70-74 (+7.51 points) and 62-69 (+6.35 points). Financial well-being improves with age, perhaps due to accumulated resources and experience.

## Income and Education
Income shows the strongest positive link to financial well-being. People earning $150,000+ score 12.60 points higher than those earning under $20,000, with steady increases across income levels. Education also relates positively: graduate degree holders score 2.92 points higher than those without high school completion.

## Health and Psychological Distress
Better health connects to higher financial well-being, while psychological distress shows the strongest negative relationship. People with excellent health score 7.70 points higher than those with poor health. High psychological distress relates to a 16.52-point decrease in financial well-being.

## Financial Knowledge and Skills
Our supplementary model in Table 2 shows varying effects based on different measurement scales:

FSscore shows a positive relationship ($\beta$ = 0.501), measured on a normal distribution across the population.

KHscore has a larger coefficient ($\beta$ = 3.748), measured across four categories (-2 to 1).

LMscore shows a moderate effect ($\beta$ = 0.874), with most people scoring the maximum value of 3.

These measurements each relate positively to financial well-being, but their different scales affect interpretation. FSscore provides the clearest measurement given its normal distribution. The effects of KHscore and LMscore need consideration alongside their measurement patterns. Better financial knowledge measures might help future studies capture population differences more precisely.

@@

## Overview
This paper examines how psychological elements, especially mental distress, work with socioeconomic factors to shape financial well-being. Our analysis of the CFPB's National Financial Well-Being Survey shows that psychological distress affects financial well-being more strongly than income, education, or financial knowledge. This finding suggests we should broaden our approach beyond financial education alone.

## Main Findings About Financial Well-Being

### Psychological Factors' Role
Psychological distress strongly influences financial well-being. Each point increase in psychological distress links to a 4.89-point decrease in financial well-being—three times the effect of one year of education. Mental health support might matter as much as financial education for improving financial outcomes.

### Financial Knowledge's Effect
Financial knowledge shows smaller effects on financial well-being than psychological and demographic factors. While financial knowledge helps, it alone may not ensure financial well-being, especially when people face psychological challenges.

## Study Limits
Our study has several constraints. The data captures one point in time, so we cannot determine cause and effect. The connection between psychological distress and financial well-being likely works both ways—financial problems may increase distress, while distress may affect financial decisions.

Self-reported data on psychological distress and financial well-being may contain biases. People with financial difficulties might report more psychological distress.

While our sample represents the nation broadly, it might miss some groups' experiences, such as people without bank accounts or internet access.

## Policy Suggestions and Next Steps
These findings point to several recommendations:

1. Financial wellness programs should include mental health components or screening.

2. Financial education initiatives should expand beyond teaching money concepts to include stress management skills.

3. Financial professionals should learn to recognize mental distress signs and connect people with mental health resources.

Future research should:
- Track people over time to understand how psychological distress affects financial outcomes
- Test programs that combine financial education with mental health support
- Study how psychological distress affects different groups' financial well-being

In summary, improving financial well-being requires attention to both financial knowledge and psychological health. We should develop approaches that address financial and psychological aspects of money management.

@@

The CFPB Financial Well-Being Scale methodology warrants examination in three areas: sampling design, measurement validation, and methods.

### Sampling Framework and Implementation
The survey collected data through multiple rounds using a national survey provider. The first round tested 46 potential items with 4,500 respondents, split between ages 18-61 and 62+. This sample matched U.S. Census demographics for working-age and older adults, though it did not aim for full national representation. The survey included both self-administered and interviewer-administered formats to test how different groups responded to the questions.

The sampling had some limits. It may not fully represent people who move frequently or are hard to reach. The removal of 500 paper surveys due to data quality concerns shows differences between collection methods that might affect how widely the findings apply.

### Measurement Development and Validation
The Financial Well-Being Scale development used item response theory (IRT), refining many initial items to create the final set. Analysis by Vector Psychometric Group supported using a bi-factor model that measures overall financial well-being while accounting for how questions were worded.

The scale showed strong internal consistency with 0.89 reliability, and the short version maintained 0.80 reliability. However, the study did not measure how stable scores remained over time. The small group of 19 people in cognitive interviews may not capture enough diverse perspectives.

Testing across demographic groups and survey types showed consistent scores in different settings. The scale distinguished between groups as expected and aligned with objective financial measures. However, limited testing with other datasets restricts how broadly we can apply these results.

### Methods: Strengths and Limits
Using different survey formats helped assess how response methods affect results. Quality checks monitored attention and completion time, though the specific standards used remain unclear.

Key limitations:

1. Some groups may be underrepresented despite address-based sampling

2. Self-reported financial data may contain bias

3. We cannot tell if financial well-being scores stay stable over time

While response rates exceeded typical online surveys, we need more analysis of who did not respond. Multiple contact attempts and payments encouraged participation, but questions about response patterns remain.

The methods balance practical needs with survey design standards, following current research practices while noting common limitations in financial studies.
